Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-08-18 paper.tex
Comments up to Section V

The marginalized PDF (Eq.~\eqref{eq:margEOS}) is now
\begin{equation}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) = &\frac{1}{p(D | \mathcal{H},\mathcal{I})} 
\int d\vec\theta_{\rm in, 1} \dots d\vec\theta_{\rm in, n} \left\{ p(\vec E | \mathcal{H},\mathcal{I}) \right. \\
& \times \prod_{i=1}^n \left[p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left.\left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) \right]\right\},
\end{split}
\end{equation}
where we have defined the quasilikelihood for the intrinsic parameters as
\begin{equation}
\label{eq:quasilike}
L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) = \int d\vec\theta_{\rm ex, i} p(\vec\theta_{\rm ex, i} | \mathcal{H}, \mathcal{I}) p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I}).
\end{equation}
Because $\tilde\Lambda_i$ is a deterministic function of $m_{1i}$, $m_{2i}$ and the EOS parameters,
\begin{equation}
p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) = \delta(\tilde\Lambda_i - \tilde\Lambda(m_{1i}, m_{2i}, \vec E)).
\end{equation}
The marginalized PDF finally becomes
\begin{equation}
\label{eq:margEOS2}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) =& \frac{1}{p(D | \mathcal{H},\mathcal{I})} \int dm_{11} dm_{21} \dots dm_{1n} dm_{2n} \\
&\times p(\vec E | \mathcal{H},\mathcal{I}) \prod_{i=1}^n \left[ p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I})|_{\tilde\Lambda_i = \tilde\Lambda(m_{1i}, m_{2i}, \vec E)} \right].
\end{split}
\end{equation}
The problem has now been reduced to computing the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each GW event, then computing Eq.~\eqref{eq:margEOS2}.

@@

The final ingredient we need to evaluate the marginalized PDF is an expression for the likelihood $p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I})$ for each GW event\footnote{In this subsection, which discusses the likelihood for individual GW events, we omit the event index $i$ for clarity.}. In this paper we assume that each detector in the network has stationary, Gaussian noise and that the noise between detectors is uncorrelated. This means that the power spectral density (PSD) $S_n(f)^a$ of the noise $n^a(t)$ in detector $a$ is \red{[[cite]]}
\begin{equation}
\langle \tilde n^a(f) \tilde n^{b*}(f') \rangle = \frac{1}{2} \delta^{ab} \delta (f-f') S_n(f)^a,
\end{equation}
where $\tilde n^a(f)$ is the Fourier transform of the noise of detector $a$, and $\langle\cdot\rangle$ represents an ensemble average. For a GW event with true parameters $\hat\theta$, resulting in the gravitational wave signal $h^a(t; \hat\theta)$, the data stream of detector $a$ will be
\begin{equation}
d^a(t) = n^a(t) + h^a(t; \hat\theta).
\end{equation}

@@

We evaluate Eq.~\eqref{eq:margEOS2} for the marginalized PDF in a two-step procedure, similar to that described by Steiner {\it et al.}~\cite{SteinerLattimerBrown2010}, where we first evaluate the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each of the $n$ BNS systems, then evaluate Eq.~\eqref{eq:margEOS2}. In the first step where we evaluate the quasilikelihood $L(\vec d; \vec\theta_{\rm in}, \mathcal{H}, \mathcal{I})$, we use the MCMC sampler \texttt{lalinference\_mcmc} included in the LSC Algorithm Library~\cite{lal} as was implemented in Ref.~\cite{WadeCreightonOchsner2014} and is described in more detail in Ref.~\cite{VanDerSluysRaymondMandel2008}. We use a uniform prior distribution for $\log(D_L)$, $\alpha$, $\sin(\delta)$, $\psi$, $\sin(\iota)$, $t_c$, and $\phi_c$ for the extrinsic parameters. \red{[[The specific range is blah....]]} The priors for the intrinsic parameters of interest $(m_{1}, m_{2}, \tilde\Lambda)$ are not contained in the quasilikelihood (Eq.~\eqref{eq:quasilike}), so we are required to use flat priors. In most cases we used $1M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 3000$. However, when we injected BNS systems with component masses of $1M_\odot$ in Section~\ref{sec:varymass}, we used $0.5M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 5000$ so that the posterior was minimally affected by the prior. For $\delta\tilde\Lambda$, we used the range $-500 \le \delta\tilde\Lambda \le 500$\red{[[Is this true?]]}. For each binary event we perform 10 independent runs and test for convergence using the Gelman-Rubin diagnostic~\cite{GelmanRubin1992}, then join the samples. Marginalization over the extrinsic parameters is performed by generating a distribution from the samples of the three intrinsic parameters with a Gaussian kernel density estimator.

In the second step, we sample the $4+2n$ parameter integrand of Eq.~\eqref{eq:margEOS2} using the affine-invarient ensemble sampler \texttt{emcee}~\cite{ForemanMackeyHogg2013}. We use uniform priors for the EOS parameters with boundaries as described in Section~\ref{sec:eosparam} and additional boundaries from the $1.93M_\odot$ observation and causality constraints as described in Sections~\ref{sec:eosconstraints} and~\ref{sec:eosparam}, and uniform priors for the masses with $1M_\odot \le m_2 \le m_1 \le 3.0M_\odot$ or, when we examine $1M_\odot$ systems, $0.5M_\odot \le m_2 \le m_1 \le 3.0M_\odot$. We again perform 10 runs and test for convergence with the Gelman-Rubin diagnostic before joining the samples. The marginalized distribution $p(\vec E|D, \mathcal{H}, \mathcal{I})$ is then given by a histogram of the samples for $\vec E$.

>>>>

The marginalized PDF (Eq.~\eqref{eq:margEOS}) is now \les{(Subsequent equation runs into next column.)}{}
\begin{equation}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) = &\frac{1}{p(D | \mathcal{H},\mathcal{I})} 
\int d\vec\theta_{\rm in, 1} \dots d\vec\theta_{\rm in, n} \left\{ p(\vec E | \mathcal{H},\mathcal{I}) \right. \\
& \times \prod_{i=1}^n \left[p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left.\left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) \right]\right\},
\end{split}
\end{equation}
where we have defined the quasilikelihood for the intrinsic parameters as
\begin{equation}
\label{eq:quasilike}
L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) = \int d\vec\theta_{\rm ex, i} p(\vec\theta_{\rm ex, i} | \mathcal{H}, \mathcal{I}) p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I}).
\end{equation}
Because $\tilde\Lambda_i$ is a deterministic function of $m_{1i}$, $m_{2i}$ and the EOS parameters,
\begin{equation}
p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) = \delta(\tilde\Lambda_i - \tilde\Lambda(m_{1i}, m_{2i}, \vec E)).
\end{equation}
The marginalized PDF finally becomes
\begin{equation}
\label{eq:margEOS2}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) =& \frac{1}{p(D | \mathcal{H},\mathcal{I})} \int dm_{11} dm_{21} \dots dm_{1n} dm_{2n} \\
&\times p(\vec E | \mathcal{H},\mathcal{I}) \prod_{i=1}^n \left[ p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I})|_{\tilde\Lambda_i = \tilde\Lambda(m_{1i}, m_{2i}, \vec E)} \right].
\end{split}
\end{equation}
The problem has now been reduced to computing the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each GW event, then computing Eq.~\eqref{eq:margEOS2}.

@@

The final ingredient we need to evaluate the marginalized PDF is an expression for the likelihood $p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I})$ for each GW event\footnote{In this subsection, which discusses the likelihood for individual GW events, we omit the event index $i$ for clarity.}. In this paper we assume that each detector in the network has stationary, Gaussian noise and that the noise between detectors is uncorrelated. This means that the power spectral density (PSD) $S_n(f)^a$ of the noise $n^a(t)$ in detector $a$ is \red{[[cite]]}
\begin{equation}
\langle \tilde n^a(f) \tilde n^{b*}(f') \rangle = \frac{1}{2} \delta^{ab} \delta (f-f') S_n(f)^a,
\end{equation}
\les{(I don't understand this notation.  If the letter $a$ labels a detector, then $\tilde{n}^{b*}$ would be the noise in detector $b$, right?)}{} where $\tilde n^a(f)$ is the Fourier transform of the noise of detector $a$, and $\langle\cdot\rangle$ represents an ensemble average. For a GW event with true parameters $\hat\theta$, resulting in the gravitational wave signal $h^a(t; \hat\theta)$, the data stream of detector $a$ will be
\begin{equation}
d^a(t) = n^a(t) + h^a(t; \hat\theta).
\end{equation}

@@

We evaluate Eq.~\eqref{eq:margEOS2} for the marginalized PDF in a two-step procedure, similar to that described by Steiner {\it et al.}~\cite{SteinerLattimerBrown2010}, where we first evaluate the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each of the $n$ BNS systems, then evaluate Eq.~\eqref{eq:margEOS2}. In the first step where we evaluate the quasilikelihood $L(\vec d; \vec\theta_{\rm in}, \mathcal{H}, \mathcal{I})$, we use the MCMC sampler \texttt{lalinference\_mcmc} included in the LSC Algorithm Library~\cite{lal} as was implemented in Ref.~\cite{WadeCreightonOchsner2014} and is described in more detail in Ref.~\cite{VanDerSluysRaymondMandel2008}. We use a uniform prior distribution for $\log(D_L)$ \les{(D=[1,$\sim$820], where ~820 is the distance at which an optimally oriented 1.4:1.4 M$_\odot$ binary is detected with SNR=8/$\sqrt{3}$ in a single detector.)}{}, $\alpha$ \les{($\alpha$=[0,2$\pi$])}{}, $\sin(\delta)$ \les{($\delta$=[-$\pi$/2,$\pi$/2])}{}, $\psi$ \les{($\psi$=[0,$\pi$])}{}, $\sin(\iota)$ \les{($\iota$=[0,$\pi$])}{}, $t_c$ \les{(uniform over data segment)}{}, and $\phi_c$ \les{($\phi_{\rm c}$=[0,2$\pi$])}{} for the extrinsic parameters. \red{[[The specific range is blah....]]} The priors for the intrinsic parameters of interest $(m_{1}, m_{2}, \tilde\Lambda)$ are not contained in the quasilikelihood (Eq.~\eqref{eq:quasilike}), so we are required to use flat priors. In most cases we used $1M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 3000$. However, when we injected BNS systems with component masses of $1M_\odot$ in Section~\ref{sec:varymass}, we used $0.5M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 5000$ so that the posterior was minimally affected by the prior. For $\delta\tilde\Lambda$, we used the range $-500 \le \delta\tilde\Lambda \le 500$\red{[[Is this true?]]}\les{(Yes.)}{}. For each binary event we perform 10 independent runs \les{(This isn't true.  I performed 5 independent runs when injecting into real noise, and 2 independent runs when injecting into zero-noise.)}{} and test for convergence using the Gelman-Rubin diagnostic~\cite{GelmanRubin1992}, then join the samples. Marginalization over the extrinsic parameters is performed by generating a distribution from the samples of the three intrinsic parameters with a Gaussian kernel density estimator.

In the second step, we sample the $4+2n$ parameter integrand of Eq.~\eqref{eq:margEOS2} using the affine-invarient ensemble sampler \texttt{emcee}~\cite{ForemanMackeyHogg2013}. We use uniform priors for the EOS parameters with boundaries as described in Section~\ref{sec:eosparam} and additional boundaries from the $1.93M_\odot$ observation and causality constraints as described in Sections~\ref{sec:eosconstraints} and~\ref{sec:eosparam}, and uniform priors for the masses with $1M_\odot \le m_2 \le m_1 \le 3.0M_\odot$ or, when we examine $1M_\odot$ systems, $0.5M_\odot \le m_2 \le m_1 \le 3.0M_\odot$ \les{(Above, the mass priors go up to 30 M$_\odot$, but here they go to just 3.  This is fine, of course, but you might want to mention that this is okay because there's no likelihood/posterior support above this region.)}{}. We again perform 10 runs and test for convergence with the Gelman-Rubin diagnostic before joining the samples. The marginalized distribution $p(\vec E|D, \mathcal{H}, \mathcal{I})$ is then given by a histogram of the samples for $\vec E$.
